### 2010-04-02 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: suppress line break before equations: we are at 11.5 pages
Desc@ICFP: s/~/\:/ in equations
Desc@ICFP: uniformize titles.
Desc@ICFP: uniformize coding style

\section{Levitating the universe of descriptions}
\label{sec:desc-levitate}

In this section, we will fulfil our promises. We first implement
finite sets. Then, we implement the codes of the $\Desc$ universe.
Despite being a perilous pedagogical exercise for the authors, we hope
to convey to the reader the dizzy feeling of levitation, without the
falling.

\subsection{Implementing finite sets}

@@

In Section~\ref{sec:finite-sets}, we have given a specification of
finite sets. We are going to implement the $\EnumU$ type former and
its constructors. Let us recall their specification:

\[\stkc{
%% EnumU
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Type{\EnumU}} 
\\
%% NilE
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Bhab{\NilE}{\EnumU}} 
\qquad
%% ConsE
\Rule{\Gamma \vdash \Bhab{\V{t}}{\UId} \quad
      \Gamma \vdash \Bhab{\V{e}}{\EnumU}}
     {\Gamma \vdash \Bhab{\ConsE{\V{t}}{\V{e}}}{\EnumU}}
}\]

The names $\NilE$ and $\ConsE{\!}{\!}$ are not a coincidence. $\EnumU$
is indeed isomorphic to a list of $\UId$s. Therefore, the actual
implementation of $\EnumU$ boils down to the following definition:

\[\stk{
\EnumU : \Set \\
\EnumU \mapsto \Mu{(\ListD~\UId)}
}\]

@@

Let us consider the consequences. First, we discover that the type
theory does not need to be extended with the type former $\EnumU$, the
constructors $\NilE$ and $\ConsE$. $\EnumU$ is defined above, as the
fixpoint of $\ListD~\UId$. $\NilE$ and $\ConsE$ are simply the type
constructors of lists, $\ListNil$ and $\ListCons$.

Another interesting consequence concerns the $\spi{\!}{\!}$ operator:
it can actually be implemented with the generic $\F{induction}$
principle. From an implementation perspective, this means that this
operator is implemented in the dependently-typed target language, as
any other function.

@@

In this section, we have replaced a low-level presentation of finite
sets by a self-hosted one, expressed in the universe of
descriptions. However, formally, the content of the type theory
remains unchanged: objects that were present before the modification
are still there. Conversely, we have not introduced any spurious
object.

If not on the content, this modification had an effect on the
names. The type naming scheme of the type theory has condensed: named
type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
now replaced by codes and their fixpoint in the universe of
descriptions. In essence, our next step is similar: we are going to
condense the entire naming scheme of datatypes \emph{in itself}.

@@

We shall now fulfil our promise concerning $\Desc$. We are going to
implement the codes of the universe of descriptions. In and by itself,
the codes, $\Desc$, is nothing but a datatype. We are in the same
situation as with $\EnumU$: we ought to be able to describe the codes
of $\Desc$ in $\Desc$ itself. Hence, this code would be a first-class
citizen, born with the standard, generic equipment of datatypes.

@@

Our first attempt is the following:

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
                               & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
                                   \DIndx{\DUnit}                        \\
                                 \end{array}
                                 \right]
\end{array}
}\]

Let us explain how we have proceeded and the obstacle we face. Much as
the datatypes we have seen so far, we first choose a constructor among
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
that the tagged notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are defining.

For $\DUnit$, we immediately reach the end of the description. For
$\DIndx{\!}$, there is a single recursive argument. Describing
$\DSigma{\!}{\!}$ is problematic. Recall the specification of
$\DSigma{\!}{\!}$:

\[    \DSigma{\!}{\!} : \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc      \]

So, we first pack an element $\V{S}$ of $\Set$. Then, we would like to
express a notion of recursive argument \emph{indexed} by $\V{S}$. Because
our presentation is entirely first-order so far, we are not able to
express this notion. We shall remedy to this situation.

@@

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. Intuitively,
$\DHindx{\!}{\!}$ uses the elements of $H$ to index as many recursive
arguments:

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DHindx{\!}{\!} & \PI{\V{H}}{\Set} \Desc \To \Desc \\
    \ldots          & \:\:\ldots
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
\ldots                                    &  \ldots 
\end{array}
}
\end{array}
}\]

@@

We are now able to describe our universe of datatypes:

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!}, \DHindx{\!}{\!} ]}) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                            \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{\V{S}}{\DUnit})}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
                                   \end{array}
                             \right]
\end{array}
}\]

The $\DUnit$ and $\DIndx{\!}$ cases remain unchanged, as expected. We
successfully describe the $\DSigma{\!}{}$ case, by a simple appeal to
the higher-order induction on $\V{S}$. The $\DHindx{\!}{\!}$ case
consists in packing an $\V{H}$ in $\Set$ with a recursive argument.

@@

What happens if we unfold the definition? We ought to build the following term:

\[
\PLAM{\V{x}}{(\EnumT{\V{E}})} \switch{\V{E}}{(\LAM{\_} \Mu{\DescD})}{\V{\pi^f}}{\V{x}}
\]

But this is quite problematic. We are still in the process of
constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
begging for this very same $\DescD$. Despite our willingness, we
cannot materialise such motive. However, we perfectly know what the
motive is. Consequently, we extend the type theory with a
special-purpose operator:

\[
\begin{array}{@{}ll}
%% switchD
\F{switchD} : & \PITEL{\V{e}}{\EnumU}               
                \PITEL{\V{b}}{\spi{\V{e}}{\LAM{\_} \Desc}}
                \PITEL{\V{x}}{\EnumT{\V{e}}} \To \Desc
\end{array}
\]

The entire work of the magician stands here, in this extension. Note
that the developer do not have to chose between the two variants of
$\F{switch}$. This is dealt with by the type propagation system
(Fig.~\ref{fig:type-checking}), extended with the following inference
rule:

\[
\Rule{\Gamma \Vdash \propag{\push{\V{t}}{\spiD{\V{e}}}}
                           {\V{t'}}}
     {\Gamma \Vdash \propag{\push{\V{t}}{\EnumT{\V{e}} \To \Desc}}
                           {\PLAM{\V{x}}{(\EnumT{\V{e}})} \switchD{\V{e}}{\V{t'}}{\V{x}}}
     }\;\mbox{$\V{t}$ is $[]$ or $[\_,\_]$}
\]

@@

The catamorphism is defined by induction on the description $\V{D}$,
with a non-dependent motive targeting $\V{T}$. Given a node $\V{xs}$
and the induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:

@@

\[\stk{
\cata : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} \\
\cata\: \V{D}\: \V{T}\: \V{f\: \V{x}} \mapsto
  \F{induction}\: \V{D}\:\V{x}\: (\LAM{\_}\V{T})\: (\LAM{\V{xs}\:\V{hs}} \V{f}\: \SHED)
}\]

We are left with filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\)
at hand. Our goal is to make an element of
\(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of the right
shape, but its sub-elements are of the wrong type. On the other hand,
for each sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding
element in $\V{T}$.  Therefore, to construct an element of
\(\descop{\V{D}}{\V{T}}\), we replace the recursive components of
\(\V{xs}\) by their counterpart in \(\V{hs}\):

\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
    \Void                                                                                                         \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
    \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto       \\
\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                     \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{xs}}{}\: \pair{\V{g}}{\V{ys}}{} \mapsto   \\
\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{g}\: \V{ys}}{}
}\]

Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\:
(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the problem. In the
type theory, we have built a generic catamorphism. Any datatype will
now come equipped with this operation, for free.

@@

\subsection{The generic Free Monad}
\label{sec:desc-free-monad}

@@

Let us recall the free monad construction. Given a functor $F$, the
free monad over $F$ is defined by the following datatype:

\[
\stk{
\data \FMFreeMonad\: \PITEL{\V{F}}{\Set \To \Set} 
                     \PITEL{\V{X}}{\Set} :
                     \Set 
\where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FMVar           & \V{X} \To \FMFreeMonad\: \V{F}\: \V{X}                            \\
    \FMComposite     & \V{F} (\FMFreeMonad\: \V{F}\: \V{X}) \To \FMFreeMonad\: \V{F}\: \V{X}    
\end{array}
}
\]

Being an inductive type, this $\FMFreeMonad$ datatype is itself
defined by a pattern functor. It is given by:

\[
\FMFreeMonadD\: \V{F}\: \V{X}\: \V{Z} \mapsto \V{X} \mathop{\D{+}} \V{F} \V{Z}
\]

@@

In our setting, the free monad construction will take the functor as a
tagged description, a set $\V{X}$ of variables, and will compute the
tagged description of the corresponding free monad. Implementing this
function is surprisingly easy:

\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
    \pair{\ListCons{\DVar{}}{\V{E}}}
         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
}\]

We simply add a constructor, $\DVar{\!}$, and define its argument to
be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
$\V{E}$ and $\V{D}$ as they were, hence leaving the functor
unchanged. Unfolding the interpretation of this definition, we
convince ourselves that this corresponds to the functor
$\FMFreeMonadD$. The fixpoint operation ties the knot and gives us
the full-blown free monad construction.

@@

Our implementation will appeal to the $\cata$ function developed
previously. So, let us write down the types, and fill as much
arguments to $\cata$ as possible:

@@

\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} \\
         & \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{x}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{\FreeMonad{\V{D}}{\V{X}}})\: 
          (\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})})\: 
          \SHED\: 
          \V{x}
}\]

We are left with implementing the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
corresponds to the following definition:

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To
           \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\etag{t}\,}{\V{ys}}{} & \Con{\pair{\etag{t}}{\V{ys}}{}}
\end{array}
}\]

>>>>

\section{Levitating the Universe of Descriptions}
\label{sec:desc-levitate}

In this section, we will fulfil our promises and show how we implement
the signatures, first for the enumerations, and then for the codes of
the $\Desc$ universe.  Persuading this to perform was a perilous
pedagogical peregrination for the protagonist.  Our method was indeed to
hardwire constants implementing the signatures specified above, in the
first instance, but then attempt to replace them, step by step, with
\emph{definitions}: ``Is \(2+2\) still \(4\)?'', ``No, it's a loop!''.
But we did find a way, so now we hope to convey to the reader the
dizzy feeling of levitation, without the falling.

\subsection{Implementing finite enumerations}

@@

In Section~\ref{sec:finite-sets}, we specified the
finite sets of tags. We are going to implement the $\EnumU$ type former and
its constructors. Recall:
%
\[\Type{\EnumU}\qquad\Bhab{\NilE}{\EnumU}\qquad
\Bhab{\ConsE{(\Bhab{\M{t}}{\UId})}{(\Bhab{\M{E}}{\EnumU})}}{\EnumU}
\]
%
The $\NilE$ and $\ConsE{\!}{\!}$ constructors are just the `nil' and
`cons' or ordinary lists, with elements from \(\UId\).
Therefore, we
implement:
%
\[
\EnumU \mapsto \Mu{(\ListD\: \UId)}
\qquad
\NilE \mapsto \ListNil
\qquad
\ConsE{\M{t}}{\M{E}} \mapsto \ListCons{\M{t}}{\M{E}}
\]

@@

Let us consider the consequences. We discover that the type
theory does not need to be extended with a special type former $\EnumU$,
or special constructors $\NilE$ and $\ConsE$. 
Moreover, the $\spi{\M{E}}{\M{P}}$ operator, computing tuple types
of \(\M{P}\)s by recursion on \(\M{E}\) need not be hardwired: we can
just use the generic $\F{ind}$ operator, as we would for any ordinary
program.

Note, however, that the universe decoder \(\EnumT{\M{E}}\) \emph{is}
hardwired, as are the primitive \(\Ze\) and \(\Su{-}\) constructs that
we use for low-level values, and indeed the \(\F{switch}\) operator.
We cannot dispose of data altogether! We have, however, gained
the ordinariness of the enumeration \emph{codes}, and hence of
programs which manipulate them. Our next step is similar: we are going to
condense the entire naming scheme of datatypes \emph{into itself}.

@@

%In this section, we have replaced a low-level presentation of finite
%sets by a self-hosted one, expressed in the universe of
%descriptions. However, formally, the content of the type theory
%remains unchanged: objects that were present before the modification
%are still there. Conversely, we have not introduced any spurious
%object.

%If not on the content, this modification had an effect on the
%names. The type naming scheme of the type theory has condensed: named
%type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
%now replaced by codes and their fixpoint in the universe of
%descriptions. In essence,

@@

We shall now fulfil our implementation promises, delivering the
codes of the universe of descriptions. In and by itself,
the codes, $\Desc$, is nothing but a datatype. We are in the same
situation as with $\EnumU$: we ought to be able to describe the codes
of $\Desc$ in $\Desc$ itself. Hence, this code would be a first-class
citizen, born with the standard, generic equipment of datatypes.

@@

Our first attempt is the following:
%
\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
                               & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
                                   \DIndx{\DUnit}                        \\
                                 \end{array}
                                 \right]
\end{array}
}\]
%
Let us explain how we have proceeded and the obstacle we face. Much as
the datatypes we have seen so far, we first choose a constructor among
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
that the tagged notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are defining.

For $\DUnit$, we immediately reach the end of the description. For
$\DIndx{\!}$, there is a single recursive argument. Describing
$\DSigma{\!}{\!}$ is problematic. Recall the specification of
$\DSigma{\!}{\!}$:
%
\[
      \DSigma{\!}{\!} : \PITEL{\V{S}}{\Set}
                        \PI{\V{D}}{\V{S} \To \Desc} 
                        \Desc      
\]
%
So, we first pack an element $\V{S}$ of $\Set$. Then, we would like to
express a notion of recursive argument \emph{indexed} by $\V{S}$. Because
our presentation is entirely first-order so far, we are not able to
express this notion. We shall remedy to this situation.

@@

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. Intuitively,
$\DHindx{\!}{\!}$ uses the elements of $H$ to index as many recursive
arguments:
%
\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DHindx{\!}{\!} & \PI{\V{H}}{\Set} \Desc \To \Desc \\
    \ldots          & \:\:\ldots
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
\ldots                                    &  \ldots 
\end{array}
}
\end{array}
}\]

@@

We are now able to describe our universe of datatypes:
%
\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!}, \DHindx{\!}{\!} ]}) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                            \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{\V{S}}{\DUnit})}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
                                   \end{array}
                             \right]
\end{array}
}\]
%
The $\DUnit$ and $\DIndx{\!}$ cases remain unchanged, as expected. We
successfully describe the $\DSigma{\!}{}$ case, by a simple appeal to
the higher-order induction on $\V{S}$. The $\DHindx{\!}{\!}$ case
consists in packing an $\V{H}$ in $\Set$ with a recursive argument.

@@

What happens if we unfold the definition? We ought to build the following term:
%
\[
\PLAM{\V{x}}{(\EnumT{\V{E}})} \switch{\V{E}}{(\LAM{\_} \Mu{\DescD})}{\V{\pi^f}}{\V{x}}
\]
%
But this is quite problematic. We are still in the process of
constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
begging for this very same $\DescD$. Despite our willingness, we
cannot materialise such motive. However, we perfectly know what the
motive is. Consequently, we extend the type theory with a
special-purpose operator:
%
\[
\begin{array}{@{}ll}
%% switchD
\F{switchD} : & \PITEL{\V{e}}{\EnumU}               
                \PITEL{\V{b}}{\spi{\V{e}}{\LAM{\_} \Desc}}
                \PITEL{\V{x}}{\EnumT{\V{e}}} \To \Desc
\end{array}
\]
%
The entire work of the magician stands here, in this extension. Note
that the developer do not have to chose between the two variants of
$\F{switch}$. This is dealt with by the type propagation system
(Fig.~\ref{fig:type-checking}), extended with the following inference
rule:
%
\[
\Rule{\Gamma \Vdash \propag{\push{\V{t}}{\spiD{\V{e}}}}
                           {\V{t'}}}
     {\Gamma \Vdash \propag{\push{\V{t}}{\EnumT{\V{e}} \To \Desc}}
                           {\PLAM{\V{x}}{(\EnumT{\V{e}})} \switchD{\V{e}}{\V{t'}}{\V{x}}}
     }\;\mbox{$\V{t}$ is $[]$ or $[\_,\_]$}
\]

@@

@@

The catamorphism is defined by induction on the description $\V{D}$,
with a non-dependent motive targeting $\V{T}$. Given a node $\V{xs}$
and the induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:
%
\[\stk{
\cata : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} \\
\cata\: \V{D}\: \V{T}\: \V{f\: \V{x}} \mapsto
  \F{induction}\: \V{D}\:\V{x}\: (\LAM{\_}\V{T})\: (\LAM{\V{xs}\:\V{hs}} \V{f}\: \SHED)
}\]
%
We are left with filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\)
at hand. Our goal is to make an element of
\(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of the right
shape, but its sub-elements are of the wrong type. On the other hand,
for each sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding
element in $\V{T}$.  Therefore, to construct an element of
\(\descop{\V{D}}{\V{T}}\), we replace the recursive components of
\(\V{xs}\) by their counterpart in \(\V{hs}\):
%
\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
    \Void                                                                                                         \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}\: \V{d'}             \mapsto
    \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}\: \pair{\V{y}}{\V{d'}}{} \mapsto        \\
\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                      \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d'}}{}\: \pair{\V{g}}{\V{d'}}{} \mapsto   \\
\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
}\]
%
Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\:
(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the problem. In the
type theory, we have built a generic catamorphism. Any datatype will
now come equipped with this operation, for free.

@@

\subsection{The generic free monad}
\label{sec:desc-free-monad}

@@

Let us recall the free monad construction. Given a functor $F$, the
free monad over $F$ is defined by the following datatype:
%
\[
\stk{
\data \FMFreeMonad\: \PITEL{\V{F}}{\Set \To \Set} 
                     \PITEL{\V{X}}{\Set} :
                     \Set 
\where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FMVar           & \V{X} \To \FMFreeMonad\: \V{F}\: \V{X}                            \\
    \FMComposite     & \V{F} (\FMFreeMonad\: \V{F}\: \V{X}) \To \FMFreeMonad\: \V{F}\: \V{X}    
\end{array}
}
\]
%
Being an inductive type, this $\FMFreeMonad$ datatype is itself
defined by a pattern functor. It is given by:
%
\[
\FMFreeMonadD\: \V{F}\: \V{X}\: \V{Z} \mapsto \V{X} \mathop{\D{+}} \V{F} \V{Z}
\]

@@

In our setting, the free monad construction will take the functor as a
tagged description, a set $\V{X}$ of variables, and will compute the
tagged description of the corresponding free monad. Implementing this
function is surprisingly easy:
%
\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
    \pair{\ListCons{\DVar{}}{\V{E}}}
         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
}\]
%
We simply add a constructor, $\DVar{\!}$, and define its argument to
be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
$\V{E}$ and $\V{D}$ as they were, hence leaving the functor
unchanged. Unfolding the interpretation of this definition, we
convince ourselves that this corresponds to the functor
$\FMFreeMonadD$. The fixpoint operation ties the knot and gives us
the full-blown free monad construction.

@@

@@

Our implementation will appeal to the $\cata$ function developed
previously. So, let us write down the types, and fill as much
arguments to $\cata$ as possible:
%
\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{x} \mapsto
  \cata\: (\toDesc{\FreeMonad{\V{D}}{\V{X}}})\: 
          (\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})})\: 
          \SHED\: 
          \V{x}
}\]
%
We are left with implementing the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
corresponds to the following definition:
%
\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To
           \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\etag{t}\,}{\V{xs}}{} & \Con{\pair{\etag{t}}{\V{xs}}{}}
\end{array}
}\]
